Visualizing Independence Using Extended Association Plots

### 2005-07-18 trunk/vcd/paper/JCGS.Rnw
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@394 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@

Figure \ref{fig:goals-assoc-classic} shows the association plot for the table of
home and away goals for all 306 games in the 1995/6 season of the German soccer league
Bundesliga \citep[the data is available in the package \texttt{vcd}]{vcd:Knorr-Held:1999}.
In particular, it can be seen that there are
fewer games ending 1--0 and more ending 1--1 than would be expected under independence.
But this does not yet imply the significance of some test statistic.

\subsection{Mosaic Plots}
\citep{vcd:Hartigan+Kleiner:1981} can be seen as an extension of grouped
bar charts where width and height of the bars show the relative
frequencies of the two variables: a mosaic plot simply
consists of a collection of tiles whose
sizes are proportional to the observed cell frequencies
(see Figure \ref{fig:goals-mosaic-classic}).

@@

@@

The implementation of Friendly mosaic plots in \textsf{SAS}
uses (by default) colors in the HLS (Hue--Luminance--Saturation) color space
with hues for blue and red, full saturation, and varying luminance
for lighter colors. The current \textsf{R} implementation is
based on colors in the HSV (Hue--Saturation--Value) space where
decreasing the saturation from 1 towards 0
%% ---given the hue for blue or red and full saturation---
generates very similar colors
as increasing the luminance for HSL colors from 0.5 towards 1.
As an alternative to those HLS and HSV colors, the device-independent
HCL (Hue--Chroma--Luminance) color space \citep{vcd:Ihaka:2003} could be
used which also provides homogeneous saturations over different colors and copier
proofness. The resulting association plot (using critical values for
$\alpha = 0.1$ and $0.01$) for the Bundesliga data can be seen in
Figure \ref{fig:goals-assoc-signif}. It indicates very clearly that
there is no evidence for rejecting the hypothesis of independence
in this table. Thus, there seems to be neither an advantage for
the home team nor for the away team.

@@

@@

\nocite{*}
\bibliography{vcd}

>>>>

somewhere:
mention Augsburg stuff, visualization of log-linear models via mosaics
\citep{vcd:Theus+Lauer:1999,vcd:Hofmann:2003,vcd:Hofmann:2001}.

Association plots as residual plots for log-linear models
\citep{vcd:Meyer+Zeileis+Hornik:2003}, especially
in coplot or trellis layout.

Arthritis data \citep{vcd:Koch+Edwards:1988}

UCB admissions data \citep{vcd:Bickel+Hammel+O'Connell:1975}

@@

%% Figure \ref{fig:goals-assoc-classic} shows the association plot for the table of
%% home and away goals for all 306 games in the 1995/6 season of the German soccer league
%% Bundesliga \citep[the data is available in the package \texttt{vcd}]{vcd:Knorr-Held:1999}.
%% In particular, it can be seen that there are
%% fewer games ending 1--0 and more ending 1--1 than would be expected under independence.
%% But this does not yet imply the significance of some test statistic.

\textbf{Mosaic Plots}
\citep{vcd:Hartigan+Kleiner:1981} can be seen as an extension of grouped
bar charts where width and height of the bars show the relative
frequencies of the two variables: a mosaic plot simply
consists of a collection of tiles whose
sizes are proportional to the observed cell frequencies
(see Figure \ref{fig:goals-mosaic-classic}).

@@

currently: chosen based on HSV or HLS color spaces,
hue (red and blue) codes sign of residuals, and saturation the absolute size
in three steps, value currently unused (always set to full value)

disadvantages:
  - perceptually not uniform
    because the three perceptual dimensions (hue, lightness, saturation)
    are not properly mapped to the three dimensions of the color space
    and hence are confounded \citep{vcd:Brewer:1999}    
    in particular, saturation is not uniform across different hues.

  - can lead to color-caused optical illusions in statistical graphs
    \citep{vcd:Cleveland+McGill:1983}

  - flashy fully saturated HSV colors are good for drawing attention to a plot,
    but hard to look at for a longer time, hence mosaic displays using these 
    colors are harder to interpret

  - white is not very suitable as a neutral color, grey conveys neutrality or 
    uninterestingness much better

Alternatives:
  - \cite{vcd:Munsell:1905} introduced color notation for balanced colors
  - Commission Internationale de l'\'Eclairage (CIE) introduced two perceptually
    based color spaces CIELAB and CIELUV where the latter is preferred for 
    emissive color technologies such as computer displays.
  - \pkg{ColorBrewer.org} \citep{vcd:Harrower+Brewer:2003} is an online tool
    for selcting different sets of colors, based on similar ideas. It includes
    qualitative, sequential and diverging palettes.
  - \cite{vcd:Ihaka:2003} discusses how CIELUV colors can be used for choosing
    colors for statistical graphics such as barplots. He recommends to use
    polar coordinates in the CIELUV space---called HCL space, using hue, chroma
    and luminance---and suggests a strategy for choosing qualitative schemes based
    on these.

We discuss how similar ideas can be used for deriving a diverging HCL scheme that 
is suitable for visualizing residuals in association and mosaic displays.

HCL
  - irregular double cone, i.e., chroma/luminance plane looks rather different
    for different hues
  - choose two hues that look similar in the chroma/luminance plane to code
    the sign of the residuals. Also use red and blue at full chroma for large
    residuals and use a light gray (instead of a white at the top of the luminance
    scale) as the neutral color for small residuals, interpolate between.
  - for non-significant results, clearly reduce the amount of color, i.e. the maximum
    chroma used.

Advantages
  - based on a perceptually uniform color space
  - device independent

@@

The implementation of Friendly mosaic plots in \textsf{SAS}
uses (by default) colors in the HLS (Hue--Luminance--Saturation) color space
with hues for blue and red, full saturation, and varying luminance
for lighter colors. The current \textsf{R} implementation is
based on colors in the HSV (Hue--Saturation--Value) space where
decreasing the saturation from 1 towards 0
%% ---given the hue for blue or red and full saturation---
generates very similar colors
as increasing the luminance for HLS colors from 0.5 towards 1.
As an alternative to those HLS and HSV colors, the device-independent
HCL (Hue--Chroma--Luminance) color space \citep{vcd:Ihaka:2003} could be
used which also provides homogeneous saturations over different colors and copier
proofness. The resulting association plot (using critical values for
$\alpha = 0.1$ and $0.01$) for the Bundesliga data can be seen in
Figure \ref{fig:goals-assoc-signif}. It indicates very clearly that
there is no evidence for rejecting the hypothesis of independence
in this table. Thus, there seems to be neither an advantage for
the home team nor for the away team.

@@

\cite{vcd:Friendly:1999} discusses grouping similar to coplots (conditioning
plots) \citep{vcd:Cleveland:1993} that lead to trellis graphics
(Becker, R.A., Cleveland, W.S. and Shyu, M. ``The Visual Design and
Control of Trellis Display'', Journal of Computational and
Graphical Statistics).

@@

\bibliography{vcd}
